\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

%% --- Bibliography (biblatex + biber) ---
\usepackage[backend=biber,
            style=alphabetic,
            sorting=nyt,
            maxbibnames=6,
            backref=true]{biblatex}
\addbibresource{references.bib}

%% --- Index (single hierarchical index) ---
\usepackage{imakeidx}
\makeindex[title=Index, intoc, columns=2]

\usepackage{hyperref}

% Suppress LaTeX automatic section numbering so only the
% explicit Gradshteyn-Ryzhik numbers appear in headings.
\setcounter{secnumdepth}{-1}

% Allow page breaks between consecutive headings (no body text).
% Without this, LaTeX suppresses breaks after headings, causing
% massive vbox overflow when headings appear back-to-back.
\makeatletter
\renewcommand{\@afterheading}{}
\makeatother
\raggedbottom

\begin{document}

%% ============================================================
%% Title and Front Matter
%% ============================================================

\begin{center}
{\LARGE\bfseries The Missing Manual for Gradshteyn--Ryzhik}\\[6pt]
{\Large A Companion to \emph{Table of Integrals, Series, and Products}}\\[12pt]
{\large John Janik}\\[4pt]
{\normalsize \today}
\end{center}

\bigskip

\subsection*{Preface}

Gradshteyn and Ryzhik's \emph{Table of Integrals, Series, and Products}
\cite{GradshteynRyzhik2007}---known universally as ``G\&R''---has been an
indispensable reference for physicists, engineers, and mathematicians since
its first Russian edition in 1943.  Its roughly 12{,}000 formulas cover
finite sums, indefinite and definite integrals, special functions,
inequalities, matrices, differential equations, and integral transforms.

G\&R tells you \emph{what} an integral evaluates to.  It does not tell you
\emph{why you would ever need it.}

This companion fills that gap.  For every section of the table, it catalogs
the physics and mathematics problems where the integrals, identities, and
special functions actually appear.  The aim is threefold:

\begin{enumerate}
\item \textbf{Forward lookup.}  You know which section of G\&R you are
  reading and want to see what it is used for---which physical systems
  produce these integrals, which mathematical structures they encode.

\item \textbf{Reverse lookup.}  You know the physics
  (e.g., ``hydrogen atom,'' ``Casimir effect,'' ``Voigt profile'') and want
  to find which integrals and formulas you will need.  The index, with
  over 5{,}200 entries, is designed for this purpose.

\item \textbf{Context and connections.}  Many formulas in G\&R are related
  to one another in ways that are invisible from the table itself.  The
  Gaussian integral (3.32) is the foundation of quantum field theory.  The
  Selberg integral (4.63) generalises the beta function to random matrix
  theory.  The Dirichlet integral $\int_{0}^{\infty}\sin x/x\,dx=\pi/2$
  (3.72) is simultaneously the normalisation of the sinc function, the
  impulse response of an ideal low-pass filter, and a special case of the
  Fourier transform.  This companion makes those connections explicit.
\end{enumerate}

\subsection*{How to use this document}

Each section of the companion follows the numbering of G\&R.  Within each
section, closely related subsections are grouped together and annotated with
two blocks:

\begin{itemize}
\item \textbf{Physics applications}---concrete problems from quantum
  mechanics, electrodynamics, statistical mechanics, fluid dynamics,
  general relativity, optics, signal processing, and other fields.
\item \textbf{Mathematics applications}---connections to analysis, algebra,
  number theory, geometry, topology, and combinatorics.
\end{itemize}

\noindent
The fastest way into the document is through the \textbf{Index} (page~\pageref{theindex}
onward).  Search for a concept---``Fourier transform,'' ``Bessel functions,''
``partition function''---and follow the page references to the relevant G\&R
sections.

The document can also be read linearly as a survey of how each class of
integrals connects to research problems.  Graduate students may find it
useful as a companion to a mathematical methods course; working researchers
may find it useful when encountering an unfamiliar integral or special
function.

\subsection*{A simple example: the hydrogen atom}

To illustrate how a single physical problem threads through the table,
consider the non-relativistic hydrogen atom.  Solving the Schr\"{o}dinger
equation $-(\hbar^{2}/2m)\nabla^{2}\psi+V(r)\psi=E\psi$ for the Coulomb
potential $V=-e^{2}/(4\pi\varepsilon_{0}r)$ requires:
%
\begin{itemize}
\item the Laplacian in spherical coordinates (G\&R~10.51--10.61),
\item separation into radial and angular equations (G\&R~16),
\item the associated Legendre functions and spherical harmonics for the
  angular part (G\&R~8.81--8.85),
\item the associated Laguerre polynomials for the radial part
  (G\&R~8.97),
\item normalisation integrals $\int_{0}^{\infty}r^{2}|R_{n\ell}|^{2}\,dr=1$
  involving exponentials times powers (G\&R~3.38--3.39), which evaluate to
  gamma functions (G\&R~8.31),
\item orthogonality integrals for the Laguerre polynomials
  (G\&R~7.414),
\item dipole matrix elements for transition rates, requiring the coupling
  of spherical harmonics (Clebsch--Gordan coefficients, G\&R~8.85) and
  radial overlap integrals (G\&R~3.38, 7.41).
\end{itemize}
%
One problem, eight sections of the table.  Add the fine structure (Dirac
equation and confluent hypergeometric functions, G\&R~9.21), the Lamb shift
(QED radiative corrections and logarithmic integrals, G\&R~4.38), or the
Stark effect (parabolic coordinates and parabolic cylinder functions,
G\&R~9.24), and the thread extends further.  Every section of G\&R is
connected to problems like this.

\subsection*{License}

This work is released under the MIT License.  You are free to copy,
distribute, modify, and use this document for any purpose, including
commercial use, provided that the copyright notice and license text are
preserved.  See the accompanying \texttt{LICENSE} file for the full terms.

\bigskip
\begin{center}
\rule{0.5\textwidth}{0.4pt}
\end{center}
\bigskip

%% ============================================================
%% Per-section files
%% ============================================================

\input{sections/sec00_introduction}

\input{sections/sec01_elementary}

\input{sections/sec02_indefinite_elementary}

\input{sections/sec0304_definite_elementary}

\input{sections/sec05_indefinite_special}

\input{sections/sec0607_definite_special}

\input{sections/sec0809_special_functions}

\input{sections/sec10_vector_field}

\input{sections/sec11_algebraic_ineq}

\input{sections/sec12_integral_ineq}

\input{sections/sec13_matrices}

\input{sections/sec14_determinants}

\input{sections/sec15_norms}

\input{sections/sec16_odes}

\input{sections/sec17_transforms}

\input{sections/sec18_ztransform}

%% ============================================================
%% Back Matter
%% ============================================================

\printbibliography[heading=bibintoc, title={Bibliography}]

\label{theindex}%
\printindex

\end{document}
